The CrypTool Script: \\ Mathematics and Cryptography \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free elearnig program CrypTool (version 1.3.10)} \vspace{60pt}

### 2005-07-04 script/en/chapters/numbertheory.tex
Update - Bernhard Esslinger June 2005

\newpage
\hypertarget{Kapitel_3}{}

\section{Introduction to Elementary Number Theory with Examples}
\label{Kapitel_3}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\

@@

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorisation, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss Carl Friedrich}:}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823 $-$ Dec 29, 1891}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

Number theory has always been a source of fascination -- for both amateurs and professional mathematicians. In 
contrast to other areas of mathematics, many of the problems and theorems in number theory can be understood 
by non-experts. On the other hand, mathematicians often take a long time to find solutions to the problems or 
prove the theorems. It is therefore one thing to pose good questions but quite another matter to find the answer. One 
example of this is what is known as Fermat's Last (or large) theorem\footnote{One of the things you learn in mathematics at school is Pythagoras theorem, which states the following for a right-angle 
triangle: $a^2 + b^2 = c^2$, where $a$ and $b$ are the lengths of the sides containing the right angle and $c$ is the 
length of the hypotenuse. Fermat famously proposed that $a^n + b^n \not= c^n$ for whole-number exponents $n 
> 2$. Unfortunately, the letter in which Fermat made the claim did not have enough space for him to prove it. 
The theorem was not proved until over 300 years later \cite[p. 433-551]{Wiles1994}\index{Wiles Andrew}. }.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling Joanne} K. Rowling\footnotemark\:}\newline This isn't magic -- it's logic -- a puzzle. A 
lot of the greatest wizards haven't got an ounce of logic.}}
\end{center}

@@

This is the claim of the 1st {\em fundamental theorem of number theory} 
(= fundamental theorem of arithmetic = fundamental building block of all positive integers). This was 
formulated precisely for the first time by Carl Friedrich Gauss in his Disquisitiones Arithmeticae (1801).  
\index{Number theory!fundamental theorem}  \index{Gauss Carl Friedrich}

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{kap_3_5_2}{}
\label{Label_Kapitel_3_5_2}

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Eric Berne\footnotemark:\index{Berne Eric}}\newline
Mathematical game theory postulates players who respond rationally.
Transactional game theory, on the other hand, deals with games that are 
not rational, perhaps even {\bf irrational and thereby closer to reality}.}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{Eric Berne, ``Games People Play'', 
     rororo, (c) 1964, page 235.}
\vskip +4pt

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Kap_3_8}{}\label{patternsandstructures}
As mathematicians investigate the structure \index{Structure}
$a *x \equiv b$ mod $m$ (see \hyperlink{kap_3_5_2}
{chapter \ref{Label_Kapitel_3_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler Leonhard}}
 function $J(n)$.

@@

\hypertarget{KleinerSatzFermat-chap3}{}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat Pierre}
}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

\begin{theorem}\label{thm-zth-fermateuler}{\bf Euler-Fermat theorem (generalisation of Fermat's Little Theorem)}
For all elements $a$ in the group $\mathbb{Z}_n^*$ (i.e. $a$ and $n$ are natural numbers that are co-prime):
$$a^{J(n)} \equiv 1 {\rm ~(mod~} n). $$
\end{theorem}
\index{Euler Leonhard} \index{Fermat Pierre}

@@

The problems of \index{Factorisation} factorising\index{Factorisation!factorisation problem} $n=pq$ with $q\neq p$ and finding $J(n)$ have a similar degree of difficulty 
and if we find a solution for one of the two problems, we also have a solution for the
other\footnote{If we know the factors of $n=p*q$ with $p\neq q$, then $J(n)=(p-1)*(q-1) = n-(p+q)+1$. Additionally the factors $p$ and $q$ are
solutions of the quadratic equation
$ x^2 - (p+q) x + pq=0.
$\\
If only $n$ and $J(n)$ are known, then it is:
$pq=n$ and $p+q= n-J(n)+1.$ So you get $p$ and $q$ by solving the equation
$$ x^2 + (J(n)-n-1)x +n= 0.
$$} (please compare requisition 3 in \hyperlink{Kapitel_3_10_1}
{section \ref{Label_Kapitel_3_10_1}}).

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\pagebreak
\subsection{Multiplicative order and primitive roots}
\hypertarget{Kap_3_9}{} \label{MultOrdPrimitveRoot}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Proof of the RSA procedure with Euler-Fermat}
\hypertarget{Kapitel_3_10}{}\index{RSA}\label{rsabeweis}

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA''. However, please note that the abbreviation may be
used in other fields to mean something completely different.}$^,$\footnote{%
In literature\index{Literature} and movies\index{Movies} not only 
classical methods are used (like the secret message solved by Sherlock Holmes
in ``The Adventure of the Dancing Men'' by Arthur Conan  
Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{list}{\textbullet}{\leftmargin10pt}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999 (included in the CrypTool package
  \index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\item the novel ``Das Museum der gestohlenen Erinnerungen (Museum of
  the stolen memories)'' by Ralf Isau \index{Isau 2003}, 
  Thienemann 2003, where the last part of the oracle can only be
  solved with the joined help of the computer community.
\end{list}\vspace{-\baselineskip}\vspace{-\parskip}
}
procedure in the group $\mathbb{Z}_n^*$.

% --------------------------------------------------------------------------
\subsubsection{Basic idea of public key cryptography}
\index{Cryptography!public key}
\hypertarget{Kapitel_3_10_1}{}
\label{Label_Kapitel_3_10_1}

@@

The 2nd requirement can easily be met because there is a ``very'' large number
of prime numbers\footnote{%
According to the prime number theorem of 
Legendre\index{Legendre Adrien-Marie} and Gauss \index{Gauss Carl Friedrich} 
there are approximately $n/\ln(n)$ 
prime numbers\index{Prime number!number of} up to the number $n$. 
This means, for example, that there are $6.5*10^{74}$ prime numbers 
under $n=2^{256}$ ($=1.1*10^{77}$) and $3.2*10^{74}$ prime numbers 
under $n=2^{255}$. 
Between $2^{255}$ and $2^{256}$ there are therefore $3.3*10^{74}$ prime numbers with precisely $256$ bits. This large number is 
also the reason why we cannot simply save them all.} and because this can be ensured by a central office that issues certificates.

@@

The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem \ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source Initiative
\index{Open Source}, that one can claim patents on algorithms. But there is really
no understanding outside the U.S., that one can get a patent for a special case 
(3 factors) of an algorithm (RSA), although the patent for the general case is
almost expired.}$^,$\footnote{If the two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is not true for all $ m<n$ (although
$ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
Example: If $ n=5^2$ then according to theorem \ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$ 
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.
\item[{\bf 2.}] Select a random $e \in \{2, \cdots, n-1\}$ such that:\\
                $e$ is relatively prime to $J(n) = (p-1)*(q-1)$.
                For example, we can select $e$ such that:  
                $\max(p,q)  <  e <  J(n) - 1$\footnote{The 
                procedure also allows us to select $d$ freely and then 
                calculate $e$. However, this has practical disadvantages.
                We usually want to be able to encrypt messages ``quickly'', which 
                is why we choose a public exponent $e$ such that it has a short 
                bit length compared to the modulus $n$ and as few binary ones as 
                possible. 
                A number often used is $65537 = 2^{16}+1$, or in binary: 
                $10\cdots 0\cdots 01$.
                We want to select the publicly known $e$ to be an advantageous 
                value that allows the exponential calculation to be performed 
                quickly during encryption. The prime numbers $3, 17$ and $65537$ 
                have proved to be particularly practical for this purpose.}.
                We can then ``throw away'' $p$ and $q$.
\item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                of $e$ modulo $J(n)$\footnote{For reasons of security, 
                $d$ should not be too small.}$^,$\footnote{We start by determining either $d$ or $e$ depending 
                on the implementation.}. We can then ``throw away'' $J(n)$.
\begin{itemize}
    \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the secret key $S$ (only $d$ must be kept secret).
\item[{\bf 4.}] For encryption, the message represented as a (binary) number is divided into parts such that each part of 
                the number is less than $n$.
\item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\item[{\bf 6.}] For decryption, the cipher text represented as a binary number is divided into parts such that each part 
                of the number is less than $n$.
\item[{\bf 7.}] Decryption of the cipher text (or the parts of it) $C \in \{1, \cdots, n-1\}$:
                $$M = D ( (n, d); C ) := C^d {\rm ~(mod~} n).$$
The numbers $d, e$ and $n$ are usually extremely large (e.~g.\ $d$ and $e$ $300$ bits, $n$ $600$ bits).

@@

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Security parameters because of progress in 
               factorisation} \label{chptSecurityParam}

The complexity is basically determined by the length $l$ of the module $n$.

@@

Within the last 20 years a lot of progress has been made. Estimations about
the future development of the ability to factor RSA modules vary and depend
on some assumptions:
\begin{itemize}
  \item progression in computing performance\index{Performance} (Moore's 
      law\index{Moore's law}: every 18 month the computing power will double) 
      and in grid computing\index{Grid computing}.
  \item development of new algorithms.
\end{itemize}
Within the last years the module bit length feasible for factorisation
increased --- even without new algorithms --- by 10 bit per year. Larger
numbers require not only more time to be factored, but also huge RAM
storage for the solutions matrix being used by the best algorithms known
today.  This need for storage grows like the square root of the computation
time, i.\:e.\ also sub-exponentially. Because RAM availability increased
exponentially in the recent decades, it seems that this should not be the
limiting factor.

Within the above mentioned article \cite{Bourseau2002} Dirk Fox\footnote{%
His company Secorvo Ltd made a statement on the recommendation for key
length selection published by the GISA\index{GISA} (German Information
Security Agency). Chapter 2.3.1 of this statement contains a very
competent and understandable discussion of RSA security (this document
exists -- to my knowledge -- only in German):\\
\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
{\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}
}
prognosticates an almost linear progression with the factorisation, if all
factors are included: Each year the module length feasible for
factorisation increases by 20 bit on average.

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Further current research in factorisation and number theory}
Prime numbers are part of very many topical research areas in number theory and computer science. 
Progress made with factorisation is bigger than was estimated 5 years ago -- this
is not only due to faster computers but also new knowledge.

>>>>

\newpage
\hypertarget{Chapter_ElementaryNT}{}

\section{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003, May 2005) \\

@@

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss, Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorisation, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss, Carl Friedrich}:}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823 $-$ Dec 29, 1891}\index{Kronecker, Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind, Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

Number theory has always been a source of fascination -- for both amateurs and professional mathematicians. In 
contrast to other areas of mathematics, many of the problems and theorems in number theory can be understood 
by non-experts. On the other hand, mathematicians often take a long time to find solutions to the problems or 
prove the theorems. It is therefore one thing to pose good questions but quite another matter to find the answer. One 
example of this is what is known as Fermat's Last (or large) theorem\footnote{One of the things you learn in mathematics at school is Pythagoras theorem, which states the following for a right-angle 
triangle: $a^2 + b^2 = c^2$, where $a$ and $b$ are the lengths of the sides containing the right angle and $c$ is the 
length of the hypotenuse. Fermat famously proposed that $a^n + b^n \not= c^n$ for whole-number exponents $n 
> 2$. Unfortunately, the letter in which Fermat made the claim did not have enough space for him to prove it. 
The theorem was not proved until over 300 years later \cite[p. 433-551]{Wiles1994}\index{Wiles, Andrew}. }.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Joanne\index{Rowling,
Joanne} K. Rowling\footnotemark\:}\newline This isn't 
magic -- it's logic -- a puzzle. A lot of the greatest 
wizards haven't got an ounce of logic.}}
\end{center}

@@

This is the claim of the 1st {\em fundamental theorem of number theory} 
(= fundamental theorem of arithmetic = fundamental building block of all positive integers). This was 
formulated precisely for the first time by Carl Friedrich Gauss in his Disquisitiones Arithmeticae (1801).  
\index{Number theory!fundamental theorem}  \index{Gauss, Carl Friedrich}

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Chapter_ElementaryNT_5_2}{}
\label{Label_Chapter_ElementaryNT_5_2}

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Eric Berne\footnotemark:\index{Berne, Eric}}\newline
Mathematical game theory postulates players who respond rationally.
Transactional game theory, on the other hand, deals with games that are 
not rational, perhaps even {\bf irrational and thereby closer to reality}.}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{Eric Berne, ``Games People Play'', 
     rororo, (c) 1964, page 235.}
\vskip +4pt

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Chapter_ElementaryNT_8}{}\label{patternsandstructures}
As mathematicians investigate the structure \index{Structure}
$a *x \equiv b$ mod $m$ (see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}}
 function $J(n)$.

@@

\hypertarget{KleinerSatzFermat-chap3}{}
\begin{theorem}\label{thm-zth-fermat1}{\bf Fermat's Little Theorem}\footnote{Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat, Pierre}
}
Let $p$ be a prime number and $a$ be a random whole number, then:
$$  a^p \equiv a {\rm ~(mod~} p).$$ 
\end{theorem}
An alternative formulation of Fermat's Little Theorem is as follows:
Let $p$ be a prime number and $a$ be a random whole number that is relatively prime to $p$, then:
$$      a^{p-1} \equiv 1 {\rm ~(mod~} p).$$

\begin{theorem}\label{thm-zth-fermateuler}{\bf Euler-Fermat theorem (generalisation of Fermat's Little Theorem)}
For all elements $a$ in the group $\mathbb{Z}_n^*$ (i.e. $a$ and $n$ are natural numbers that are co-prime):
$$a^{J(n)} \equiv 1 {\rm ~(mod~} n). $$
\end{theorem}
\index{Euler, Leonhard} \index{Fermat, Pierre}

@@

The problems of \index{Factorisation} factorising\index{Factorisation!factorisation problem} $n=pq$ with $q\neq p$ and finding $J(n)$ have a similar degree of difficulty 
and if we find a solution for one of the two problems, we also have a solution for the
other\footnote{If we know the factors of $n=p*q$ with $p\neq q$, then $J(n)=(p-1)*(q-1) = n-(p+q)+1$. Additionally the factors $p$ and $q$ are
solutions of the quadratic equation
$ x^2 - (p+q) x + pq=0.
$\\
If only $n$ and $J(n)$ are known, then it is:
$pq=n$ and $p+q= n-J(n)+1.$ So you get $p$ and $q$ by solving the equation
$$ x^2 + (J(n)-n-1)x +n= 0.
$$} (please compare requisition 3 in \hyperlink{Chapter_ElementaryNT_10_1}
{section \ref{Label_Chapter_ElementaryNT_10_1}}).

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\pagebreak
\subsection{Multiplicative order and primitive roots}
\hypertarget{Chapter_ElementaryNT_9}{} \label{MultOrdPrimitveRoot}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Proof of the RSA procedure with Euler-Fermat}
\hypertarget{Chapter_ElementaryNT_10}{}\index{RSA}\label{rsabeweis}

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA''. However, please note that the abbreviation may be
used in other fields to mean something completely different.}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
(\hyperlink{appendix-movies}
{see appendix C: Movies and Literature with Relation to Cryptography}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

% --------------------------------------------------------------------------
\subsubsection{Basic idea of public key cryptography}
\index{Cryptography!public key}
\hypertarget{Chapter_ElementaryNT_10_1}{}
\label{Label_Chapter_ElementaryNT_10_1}

@@

The 2nd requirement can easily be met because there is a ``very'' large number
of prime numbers\footnote{%
According to the prime number theorem of 
Legendre\index{Legendre, Adrien-Marie} and Gauss \index{Gauss, Carl Friedrich} 
there are approximately $n/\ln(n)$ 
prime numbers\index{Prime number!number of} up to the number $n$. 
This means, for example, that there are $6.5*10^{74}$ prime numbers 
under $n=2^{256}$ ($=1.1*10^{77}$) and $3.2*10^{74}$ prime numbers 
under $n=2^{255}$. 
Between $2^{255}$ and $2^{256}$ there are therefore $3.3*10^{74}$ prime numbers with precisely $256$ bits. This large number is 
also the reason why we cannot simply save them all.} and because this can be ensured by a central office that issues certificates.

@@

The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem \ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source Initiative
\index{Open Source}, that one can claim patents on algorithms. But there is really
no understanding outside the U.S., that one can get a patent for a special case 
(3 factors) of an algorithm (RSA), although the patent for the general case is
almost expired.}$^,$\footnote{If the two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is not true for all $ m<n$ (although
$ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
Example: If $ n=5^2$ then according to theorem \ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$ 
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

\item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such that\footnote{%
                It is recommended by cryptanalytic reasons, but not necessary to
                make RSA work, to select $e$ such that: \\
                $\max(p,q)  <  e <  J(n) - 1$.}: \\
                $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                We can then ``throw away'' $p$ and $q$.\footnote{%
                The procedure also allows us to select $d$ freely and then 
                calculate $e$. However, this has practical disadvantages.
                We usually want to be able to encrypt messages ``quickly'', which 
                is why we choose a public exponent $e$ such that 
                it has a short bit length compared to the modulus $n$ and 
                as few binary ones as possible (e.g. $2^{16} + 1$).
                So a fast exponentiation is possible when encrypting.                
                We want to select the publicly known $e$ to be an advantageous 
                value that allows the exponential calculation to be performed 
                quickly during encryption. The prime numbers $3, 17$ and $65537$ 
                have proved to be particularly practical for this purpose.
                The most often used number is $65537 = 2^{16}+1$, or in binary: 
                $10\cdots 0\cdots 01$ (this number is prime and therefor relatively
                prime to many other numbers).
                }
\item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                of $e$ modulo $J(n)$\footnote{For reasons of security, 
                $d$ should not be too small.}$^,$\footnote{We start by determining either $d$ or $e$ depending 
                on the implementation.}. We can then ``throw away'' $J(n)$.
\begin{itemize}
    \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the secret key $S$ (only $d$ must be kept secret).
\end{itemize}
\item[{\bf 4.}] For encryption, the message represented as a (binary) number is divided into parts such that each part of 
                the number is less than $n$.
\item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\item[{\bf 6.}] For decryption, the cipher text represented as a binary number is divided into parts such that each part 
                of the number is less than $n$.
\item[{\bf 7.}] Decryption of the cipher text (or the parts of it) $C \in \{1, \cdots, n-1\}$:
                $$M = D ( (n, d); C ) := C^d {\rm ~(mod~} n).$$
\end{itemize}
The numbers $d, e$ and $n$ are usually extremely large (e.~g.\ $d$ and $e$ $300$ bits, $n$ $600$ bits).

@@

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Security parameters because of new algorithms} \label{chptSecurityParam}

%\vskip +15 pt
\paragraph{Factorisation algorithms} \mbox{}

The complexity is basically determined by the length $l$ of the module $n$.
Higher values for this major parameter are oriented at the possibilities of the
current algorithms for factorization:

@@

And also past 1999 further factorization progress was made up to now 
(see RSA-200 in chapter \ref{RSA-200}).

\vskip +15 pt
\paragraph{Lattice base reduction algorithms} \mbox{}

The module length is not the only parameter relevant for security.
Beneath requirements from implementation and engineering the sizes and
the proportions of the  parameters e, d and N are relevant.

According attacks based on lattice reductions\index{Lattice reduction} 
are a real threat for (too) simple implementations of RSA.
Theses attacks can be structured into the following four categories:
\begin{itemize}
   \item Attacks against very small public keys e (e.g. $e = 3$). 
   \item Attacks against relatively small private exponents d(e.g. $d < N ^ {0.5}$).
   \item Factorization of the modul N, if one of the factors p or q is {\em partly} known.
   \item Attacks requireing, that a {\em part} of the private key d is known.
\end{itemize}

A good overview about the current status of publications concerning these attacks
can be found in the diploma thesis of Matthias Schneider \cite{SchneiderM2004}.

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Forecasts about factorization of large integers}
\index{Factorisation!forecast}

Within the last 20 years a lot of progress has been made. Estimations about
the future development of the ability to factor RSA modules vary and depend
on some assumptions:
\begin{itemize}
  \item progression in computing performance\index{Performance} (Moore's 
      law\index{Moore's law}\index{Moore, Gordon E.}: every 18 month the
      computing power will double) and in grid computing\index{Grid computing}.
  \item development of new algorithms.
\end{itemize}
Within the last years the module bit length feasible for factorisation
increased --- even without new algorithms --- by 10 bit per year. Larger
numbers require not only more time to be factored, but also huge RAM
storage for the solutions matrix being used by the best algorithms known
today.  This need for storage grows like the square root of the computation
time, i.\:e.\ also sub-exponentially. Because RAM availability increased
exponentially in the recent decades, it seems that this should not be the
limiting factor.

An estimation of the evolution of secure key lengths was done by 
Lenstra/Verheul\cite{Lenstra1999} (compare figure~\ref{RSAKeylength}
in chapter ~\ref{ECAlternative}).
